### 2023-05-30 paper/localpsf.tex
edit intro

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels.
%	, with a focus on approximating high rank Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). 
%	The method may be used to approximate other operators, including certain Schur complements, Poincare-Steklov operators, covariance operators, and blurring operators. 
	The method computes impulse responses of the operator at a collection of scattered points, and interpolates translated and scaled versions of these impulse responses to approximate entries of the integral kernel. We compute a large number of impulse responses by applying the operator to a small number Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we compute as many impulse responses as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. Support ellipsoids are estimated a-priori based on impulse response moments, which are computed by applying the operator to a small number of polynomial functions. The interpolation uses a ``local mean displacement invariance'' approximation, which improves upon classical local translation invariance. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are performed with fast H-matrix methods. We apply the method to approximate the Hessians in an ice sheet flow PDE constrained inverse problem, and an advection-diffusion PDE constrained inverse problem. Numerical results demonstrate that the proposed method substantially outperforms existing methods, and can approximate high rank operators using only a small number of operator applies.
\end{abstract}

@@

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows 
%one to approximate discretized matrix representations of $\Aop$ using methods that require fast access to arbitrary matrix entries. We 
us to form a hierarchical matrix (H-matrix)~\cite{BormGrasedyckHackbusch03,Hackbusch99} approximation of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure~\ref{fig:hmatrix_intro} and Appendix~\ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks~\cite{GrasedyckHackbusch03}. The ability to perform these operations permits, for example, solution of Newton linear systems in PDE constrained optimization, sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank Gaussian surrogate models that can be used for uncertainty quantification. 
%Classical H-matrix construction methods 
%(described in Appendix~\ref{sec:H_matrix_construction}) 
%require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free.

>>>>

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels.
%	, with a focus on approximating high rank Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). 
%	The method may be used to approximate other operators, including certain Schur complements, Poincare-Steklov operators, covariance operators, and blurring operators. 
	The method computes impulse responses of the operator at a collection of scattered points, and interpolates translated and scaled versions of these impulse responses to approximate entries of the integral kernel. We compute a large number of impulse responses by applying the operator to a small number Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we compute as many impulse responses as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. Support ellipsoids are estimated a-priori based on impulse response moments, which are computed by applying the operator to a small number of polynomial functions. The interpolation uses a ``local mean displacement invariance'' approximation, which improves upon classical local translation invariance. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are performed with H-matrix methods. We apply the method to approximate the Hessians in an ice sheet flow PDE constrained inverse problem, and an advection-diffusion PDE constrained inverse problem. Numerical results demonstrate that the proposed method outperforms existing methods, and can approximate high rank operators using only a small number of operator applies.
\end{abstract}

@@

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows one to approximate discretized versions of $\Aop$ using the full arsenal of tools for matrix approximation that rely on fast access to matrix entries.
%one to approximate discretized matrix representations of $\Aop$ using methods that require fast access to arbitrary matrix entries. We 
We form a hierarchical matrix (H-matrix)~\cite{BormGrasedyckHackbusch03,Hackbusch99} approximation of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure~\ref{fig:hmatrix_intro} and Appendix~\ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks~\cite{GrasedyckHackbusch03}. The ability to perform these operations permits, for example, solution of Newton linear systems in PDE constrained optimization, sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank Gaussian surrogate models that can be used for uncertainty quantification. 
%Classical H-matrix construction methods 
%(described in Appendix~\ref{sec:H_matrix_construction}) 
%require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free.
